The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-04-14 paper.tex
Improve writing and consistently use present tense in introduction.

The science of anthropology can benefit from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
Currently problems of access could be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
The allocation of researchers and research material would greatly enhance the quality and extent of interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and adequately formalised would enable the application of efficient data analysing methods such as data mining and textmining techniques that could promote deeper insights.
Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) proposed here, is an appropriate formalisation~\citep{aw3dunpublished}.
Apart from contextual information, most informative clues can be found directly on the bone.
As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}

By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B. Beleg zur Topographie. KH: könnte man hier als Quelle die TA nehmen?}
This map would contain visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
These would be unequivocally located by accurate and exhaustive definitions that in turn would allow for their easy retrieval and moreover would serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.

@@

ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

>>>>

The science of anthropology benefits from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
The allocation of researchers and research material greatly enhances the quality and extent of interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and adequately formalised enable the application of efficient data analysing methods such as data mining and textmining techniques that promote deeper insights.
Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) presented here, is an appropriate formalisation~\citep{aw3dunpublished}.

Apart from contextual information, most informative clues can be found directly on the bone.
As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}
By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B. Beleg zur Topographie. KH: könnte man hier als Quelle die TA nehmen?}
Such a map contains visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
These are unequivocally located by accurate and exhaustive definitions that in turn allow for their easy retrieval and moreover serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.

@@

ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
